SunPy - Python for Solar Physics

### 2014-01-28 3-2-HEK.tex
Stuart's pass-the-paper

The Sun is an active star and exhibits a wide range of transient phenomena (e.g. flares, 
radio bursts) on many 
different time-scales, length-scales and wavelengths. Observations and metadata 
concerning these phenomena are collected at the Heliophysics Event 
Knowledgebase (HEK).  Entries are generated both by automated algorithms and 
human observers.  Some of the information in the HEK reproduces feature/event 
data elsewhere (for example, the GOES flare catalog), and some is generated by 
the Solar Dynamics Observatory Feature Finding Team \cite{martens2012} to
populate the HEK.  The advantage of the HEK is it provides an homogeneous and 
well-described interface to a large 
amount of feature/event information of interest to the solar physics community.

SunPy accesses this information through the \texttt{hek} module, which was 
developed with support from the 
ESA-SOCIS 2011.  The \texttt{hek} module makes use of the HEK API
(\url{http://vso.stanford.edu/hekwiki/ApplicationProgrammingInterface?action=print}).
The current list of 
events maintained by the HEK and their properties can be found at 
\url{http://www.lmsal.com/hek/VOEvent_Spec.html}.

@@

HEK attributes differ from VSO attributes (Section \ref{ssec:vso}) in that many 
of them are wrappers that conveniently expose comparisons by overloading Python 
operators.  This allows filtering of the HEK entries by the properties of the 
event.  As was mentioned above, the HEK stores feature/event metadata obtained 
in different ways, known generally as {\it feature recognition methods} ( 
FRMs). Example in Listing~\ref{code:hek:frm} filters the results of the previous result to 
return only those events that have the FRM 'SSW Latest Events'.
\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> result = client.query(hek.attrs.Time(tstart,tend), 
...                       hek.attrs.EventType(event_type),
...                       hek.attrs.FRM.Name == 'SSW Latest Events')
>>> len(result)
9
\end{minted}
\caption{An HEK query that returns only those flares that were
  detected by the 'SSW Latest Events' feature recognition method.}
\label{code:hek:frm}
\end{listing}
Multiple comparisons can be made by including more comma-separated
conditions on the attributes in the call to the HEK query method.

@@

The ability to use comparison and logical operators on HEK attributes allows 
the construction of queries of almost arbitrary complexity. 
It is also possible to use the 
results of a HEK query to download the corresponding data from the VSO by using 
SunPy's \texttt{H2VClient}.
%schriste - this is a weird way to end this section, either show how it is done or 
% do not mention it, i think.
%dps - I agree.

>>>>

The Sun is an active star and exhibits a wide range of transient phenomena 
(e.g. flares and radio bursts) on many different time-scales, length-scales and 
wavelengths. Observations and metadata concerning these phenomena are collected 
in the Heliophysics Event Knowledgebase (HEK).  Entries are generated both by 
automated algorithms and human observers.  Some of the information in the HEK 
reproduces feature/event data elsewhere (for example, the GOES flare catalogue),
and some is generated by the Solar Dynamics Observatory Feature Finding Team 
\citep{martens2012} to populate the HEK.  The advantage of the HEK is it 
provides an homogeneous and well-described interface to a large amount of 
feature/event information of interest to the solar physics community.

SunPy accesses this information through the \texttt{hek} module, which was also
developed with support from ESA's SOCIS 2011.  The \texttt{hek} module makes 
use of the 
\href{http://vso.stanford.edu/hekwiki/ApplicationProgrammingInterface?action=print}{HEK
 API}.
The current list of events maintained by the HEK and their properties can be 
found at \url{http://www.lmsal.com/hek/VOEvent_Spec.html}.

@@

HEK attributes differ from VSO attributes (Section \ref{ssec:vso}) in that many 
of them are wrappers that conveniently expose comparisons by overloading Python 
operators. This allows filtering of the HEK entries by the properties of the 
event. As was mentioned above, the HEK stores feature/event metadata obtained 
in different ways, known generally as {\it feature recognition methods} (FRMs). 
Example in Listing~\ref{code:hek:frm} filters the results of the previous 
result to return only those events that have the FRM 'SSW Latest Events'.
\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> result = client.query(hek.attrs.Time(tstart,tend), 
...                       hek.attrs.EventType(event_type),
...                       hek.attrs.FRM.Name == 'SSW Latest Events')
>>> len(result)
9
\end{minted}
\caption{An HEK query that returns only those flares that were
  detected by the 'SSW Latest Events' feature recognition method.}
\label{code:hek:frm}
\end{listing}
Multiple comparisons can be made by including more comma-separated
conditions on the attributes in the call to the HEK query method.

@@

The ability to use comparison and logical operators on HEK attributes allows 
the construction of queries of almost arbitrary complexity. 
It is also possible to use the results of a HEK query to download the 
corresponding data from the VSO by using SunPy's \texttt{H2VClient}.
%schriste - this is a weird way to end this section, either show how it is done or 
% do not mention it, i think.
%dps - I agree.
%Stuart: Will do on my next pass

### 2014-01-28 3-3-HELIO.tex
Stuart's pass-the-paper

The HELiophysics Integrated Observatory 
(HELIO, \url{http://helio-vo.eu}) has put together a list of web 
services which allows scientists to query and discover data throughout 
the heliosphere (\textit{e.g.}, solar, planetary, inter-planetary data, 
\ldots)\citep{dps2012}.
HELIO has been built with a Service-Oriented Architecture, 
\textit{i.e.} its capabilities are split into a number of tasks that are 
implemented as separate services. 
HELIO comprises nine different public services, which allows scientists
to search in different catalogues of registered events, solar features,
data from instruments in the heliosphere and other information like planets or 
spacecraft position in time. 
Additionally, HELIO provides a service that uses a propagation model to link 
the data in different points of the solar system by its original nature 
(\textit{e.g.}, Earth auroras are a signature of magnetic field disturbances 
produced few days before on the Sun).
In addition to the primary, web-based interface to HELIO
(\url{http://hfe.helio-vo.eu}), the services can be separately accessed.

SunPy's \texttt{hec} module provides an interface to the
\textit{HELIO Event Catalogue} (HEC) service, and this module was developed as
part of a Google Summer of Code project in 2013.
The HEC service currently provides access to 84 catalogues from different
sources (automated and manual classification, event lists from publications, \ldots).
As with all of the HELIO services, the HEC service provides results in VOTable 
data format (defined by IVOA \cite{ochsenbein_ivoa_2011}), and the \texttt{hec}
module parses this output using the \texttt{astropy.io.votable} package.
This format has the advantage of containing metadata with information like
data provenance and the performed query.

>>>>

The HELiophysics Integrated Observatory (\href{http://helio-vo.eu}{HELIO}) has 
put together a list of web services which allows scientists to query and 
discover data throughout the heliosphere (\textit{e.g.}, solar, planetary, 
inter-planetary data, \ldots)\citep{dps2012}.
HELIO has been built with a Service-Oriented Architecture, 
\textit{i.e.} its capabilities are split into a number of tasks that are 
implemented as separate services. 
\href{http://helio-vo.eu}{HELIO} comprises nine different public services, 
which allows scientiststo search in different catalogues of registered events, 
solar features, data from instruments in the heliosphere and other information 
like planets or spacecraft position in time. 
Additionally, \href{http://helio-vo.eu}{HELIO} provides a service that uses a 
propagation model to link the data in different points of the solar system by 
its original nature (\textit{e.g.}, Earth auroras are a signature of magnetic 
field disturbances produced few days before on the Sun).
In addition to the primary, web-based interface to 
\href{http://helio-vo.eu}{HELIO}, the services can be separately accessed.

SunPy's \texttt{hec} module provides an interface to the
\textit{HELIO Event Catalogue} (HEC) service, and this module was developed as
part of a Google Summer of Code (GSOC) project in 2013.
The HEC service currently provides access to 84 catalogues from different
sources (automated and manual classification, event lists from publications, \ldots).
As with all of the HELIO services, the HEC service provides results in VOTable 
data format (defined by IVOA \cite{ochsenbein_ivoa_2011}), and the \texttt{hec}
module parses this output using the \texttt{astropy.io.votable} package.
This format has the advantage of containing metadata with information like
data provenance and the performed query.

### 2014-01-28 4-Util.tex
Stuart's pass-the-paper

\subsection{World Coordinate System (WCS) Coordinates}\label{ssec:util:wcs}
Coordinate transformations are frequently a necessary task within the solar 
data analysis workflow. Likely the most often used transformation is from 
observer coordinates (e.g. sky coordinates) to a coordinate system which is 
mapped onto the solar surface (e.g. latitude and longitude). This 
transformation is necessary to compare the true physical distance between 
different solar features. This type of transformation is not unique
to solar observations, but is not often considered by astronomical packages
such as the Astropy 
coordinates package. The \verb|wcs| package in SunPy implements the World Coordinate 
System (WCS) for solar coordinates as described by \cite{Thompson2000}. The 
transformations currently implemented are those most useful 
for most solar data analysis, namely converting from Helioprojective-Cartesian 
(HPC) to Heliographic coordinates (HG). HPC describes the positions on 
the Sun as angles measured from the center of the solar disk (usually in 
arcseconds) using cartesian coordinates (X, Y). This is the coordinate system 
most often defined in solar imaging data (see for example, images from SDO/AIA, SOHO/EIT, and TRACE). 
HG coordinates express positions on the Sun using longitude and latitude on 
the solar sphere. There are two standards for this coordinate system, 
Stonyhurst-Heliographic where the origin is at the intersection of the solar 
equator and the central meridian as seen from Earth and 
Carrington-Heliographic is a system which is fixed to the Sun and does not depend on Earth. The 
implementation of these transformations pass through a common coordinate system 
called Heliocentric-Cartesian (HCC) where positions are expressed in true 
(deprojected) physical distances instead of angles on the celestial sphere.
These transformations require some knowledge of the location of the observer 
which is usually provided by the image header. In the case where it is 
not, the observer is assumed to be at Earth.Listing \ref{code:wcs_code} shows some examples of coordinate transforms carried out in SunPy using the \verb|wcs| utilities.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy import wcs
>>> wcs.convert_hg_hpc(10,53)
(100.49244115330731, 767.97438321917502)
# Convert that position back to heliographic coordinates
>>> wcs.convert_hpc_hg(100.49, 767.97)
(9.9996521808465175, 52.999563684874893)
# Try to convert a position which is not on the Sun to HG
>>> wcs.convert_hpc_hg(-1500,0)
sunpy/wcs/wcs.py:180: RuntimeWarning: invalid value encountered in sqrt
  distance = q - np.sqrt(distance)
(nan, nan)
Convert sky coordinate to a position in HCC
>>> wcs.convert_hpc_hcc(-300,400, z = True)
(-216716967.63331246, 288956420.9477042, 594364636.2208252)
\end{minted}
\caption{Using the sunpy.wcs sub-package.}
\label{code:wcs_code}
\end{listing}

\subsubsection{Sun}\label{ssec:util:sun}
The purpose of this module is to provide solar-specific data for example ephemeris and
solar constants. The main namespace contains a number of functions which provide solar
ephemeris such as the sun to earth distance, solar cycle number, the mean anomaly, etc.
All of these functions take a time as their input which can be provided in any 
\texttt{parse\_time}
compatible format. 
A future version of SunPy will integrate with the AstroPy coordinates package and will 
likely make use of an ephemeris package.

\subsubsection{constants}\label{ssec:util:const}
The purpose of the constants package is to provide a number of solar-related 
constants in order to provide consistency in the calculations of derived solar 
values both within the SunPy code base but also to the user. Every solar 
constant is provided as a Constant object as defined by Astropy. Each 
Constant object defines a Quantity, a number associated with a unit, along with 
the constant's provenance (i.e. reference) as well as its uncertainty. Using 
Astropy's Quantities, any solar constant can easily be converted between 
different units as can be seen in Figure~\ref{fig:constants_code} as well
as between the SI or cgs unit systems with a single command.  As these objects inherit from 
Numpy's ndarray, they work well with standard representations of numbers.
For convenience a number of shortcuts to frequently used constants are provided 
directly when importing the sub-package. A larger list of constants can be 
accessed through an interface modelled on that provided by the SciPy constants 
package and is available as a dictionary called \texttt{physical\_constants}. 
To view them all quickly a \texttt{print\_all()} function is available.

>>>>

\subsection{World Coordinate System (WCS) Coordinates}\label{ssec:util:wcs}
Coordinate transformations are frequently a necessary task within the solar 
data analysis workflow. Likely the most often used transformation is from 
observer coordinates (e.g. sky coordinates) to a coordinate system which is 
mapped onto the solar surface (e.g. latitude and longitude). This 
transformation is necessary to compare the true physical distance between 
different solar features. This type of transformation is not unique
to solar observations, but is not often considered by astronomical packages
such as the Astropy 
coordinates package. The \verb|wcs| package in SunPy implements the World Coordinate 
System (WCS) for solar coordinates as described by \cite{Thompson2000}. The 
transformations currently implemented are those most useful 
for most solar data analysis, namely converting from Helioprojective-Cartesian 
(HPC) to Heliographic coordinates (HG). HPC describes the positions on 
the Sun as angles measured from the center of the solar disk (usually in 
arcseconds) using cartesian coordinates (X, Y). This is the coordinate system 
most often defined in solar imaging data (see for example, images from 
\textit{SDO/AIA}, \textit{SOHO/EIT}, and \textit{TRACE}). 
HG coordinates express positions on the Sun using longitude and latitude on 
the solar sphere. There are two standards for this coordinate system, 
Stonyhurst-Heliographic where the origin is at the intersection of the solar 
equator and the central meridian as seen from Earth and 
Carrington-Heliographic is a system which is fixed to the Sun and does not depend on Earth. The 
implementation of these transformations pass through a common coordinate system 
called Heliocentric-Cartesian (HCC) where positions are expressed in true 
(deprojected) physical distances instead of angles on the celestial sphere.
These transformations require some knowledge of the location of the observer 
which is usually provided by the image header. In the case where it is 
not, the observer is assumed to be at Earth. Listing \ref{code:wcs_code} shows 
some examples of coordinate transforms carried out in SunPy using the 
\verb|wcs| utilities.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy import wcs
>>> wcs.convert_hg_hpc(10,53)
(100.49244115330731, 767.97438321917502)
# Convert that position back to heliographic coordinates
>>> wcs.convert_hpc_hg(100.49, 767.97)
(9.9996521808465175, 52.999563684874893)
# Try to convert a position which is not on the Sun to HG
>>> wcs.convert_hpc_hg(-1500,0)
sunpy/wcs/wcs.py:180: RuntimeWarning: invalid value encountered in sqrt
  distance = q - np.sqrt(distance)
(nan, nan)
Convert sky coordinate to a position in HCC
>>> wcs.convert_hpc_hcc(-300,400, z = True)
(-216716967.63331246, 288956420.9477042, 594364636.2208252)
\end{minted}
\caption{Using the sunpy.wcs subpackage.}
\label{code:wcs_code}
\end{listing}

\subsubsection{Sun}\label{ssec:util:sun}
The purpose of this module is to provide solar-specific data for example ephemeris and
solar constants. The main namespace contains a number of functions which provide solar
ephemeris such as the Sun to Earth distance, solar cycle number, the mean 
anomaly, etc.
All of these functions take a time as their input which can be provided in any 
\texttt{parse\_time}
compatible format. 
A future version of SunPy will integrate with the AstroPy coordinates package and will 
likely make use of an ephemeris package.

\subsubsection{constants}\label{ssec:util:const}
The purpose of the constants package is to provide a number of solar-related 
constants in order to provide consistency in the calculations of derived solar 
values both within the SunPy code base but also to the user. Every solar 
constant is provided as a Constant object as defined by Astropy. Each 
Constant object defines a Quantity, a number associated with a unit, along with 
the constant's provenance (i.e. reference) as well as its uncertainty. Using 
Astropy's Quantities, any solar constant can easily be converted between 
different units as can be seen in Listing~\ref{code:constants_code} as well
as between the SI or cgs unit systems with a single command.  As these objects inherit from 
Numpy's ndarray, they work well with standard representations of numbers.
For convenience a number of shortcuts to frequently used constants are provided 
directly when importing the sub-package. A larger list of constants can be 
accessed through an interface modelled on that provided by the SciPy constants 
package and is available as a dictionary called \texttt{physical\_constants}. 
To view them all quickly a \texttt{print\_all()} function is available.

### 2014-01-28 5-Dev.tex
Stuart's pass-the-paper

\section{Development and Community}\label{sec:dev}
SunPy is a community-developed library, designed and developed for and by 
the solar physics community. Not only is all the source code publicly available 
online under the permissive 2-clause BSD licence, the whole 
development 
process is also online and open for anyone to contribute to.
SunPy's development makes use of the online 
service GitHub (\url{http://github.com}) and Git\footnote{\url{http://git-scm.com/}}
as its distributed version control software.

@@

Quality documentation is
one of the most important factors determining the success of any software project. 
Thankfully powerful tools already exist in Python to support this thanks to native
Python's focus on its own documentation. SunPy makes use of the Sphinx (\url{http://sphinx-doc.org})
documentation generator. Sphinx uses reStructuredText as its markup language: 
an easy-to-read, what-you-see-is-what-you-get plaintext markup syntax. It supports
many output formats most notably HTML as well as PDF, and ePub and provides a rich
hierarchically structured view of in-code documentation strings. The SunPy documentation 
is built automatically and hosted on Read the Docs (\url{http://readthedocs.org})
at \url{http://docs.sunpy.org}.

@@

Many of the larger features in SunPy have been developed 
with the support of both the ESA-SOCIS and GSOC. 
Initial development of SunPy's VSO and HEK 
implementations were funded by the SOCIS program as well as a prototyped GUI 
and N-Dimensional datatype implementations. SunPy was supported by GSOC in 2013 
through the Python Software Foundation (PSF). It supported development of the HELIO and 
database sub-packages.

\subsection{Development Workflow}
The development workflow that GitHub enables is simple and transparent. The 
primary source code repository \url{http://github.com/sunpy/sunpy} can be 
`forked' by anyone with a GitHub account who can then modify their fork as they 
wish. Upon completion of any modifications it is possible for a user to create 
a 'Pull Request' (\url{http://github.com/sunpy/sunpy/pulls}) to the main SunPy 
repository in order to incorporate their modification into the SunPy code base. 
Pull requests are reviewed by the SunPy community before being accepted. Community consensus
is required to add new features.

>>>>

\section{Development and Community}\label{sec:dev}
SunPy is a community-developed library, designed and developed for and by 
the solar physics community. Not only is all the source code publicly available 
online under the permissive 2-clause BSD licence, the whole 
development process is also online and open for anyone to contribute to.
SunPy's development makes use of the online service 
\href{http://github.com}{GitHub} and \href{http://git-scm.com/}{Git}
as its distributed version control software.

@@

Quality documentation is
one of the most important factors determining the success of any software project. 
Thankfully powerful tools already exist in Python to support this thanks to native
Python's focus on its own documentation. SunPy makes use of the Sphinx (\url{http://sphinx-doc.org})
documentation generator. Sphinx uses reStructuredText as its markup language: 
an easy-to-read, what-you-see-is-what-you-get plaintext markup syntax. It supports
many output formats most notably HTML as well as PDF, and ePub and provides a rich
hierarchically structured view of in-code documentation strings. The SunPy documentation 
is built automatically and hosted by \href{http://readthedocs.org}{Read the 
Docs} at \url{http://docs.sunpy.org}.

@@

Many of the larger features in SunPy have been developed 
with the support of both the ESA-SOCIS and GSOC. 
Initial development of SunPy's VSO and HEK 
implementations were funded by the SOCIS program as well as a prototyped GUI 
and N-Dimensional datatype implementations. SunPy was supported by GSOC in 2013 
through the Python Software Foundation (PSF). It supported development of the 
\texttt{HELIO} and 
\texttt{database} subpackages.

\subsection{Development Workflow}
The development workflow that GitHub enables is simple and transparent. The 
primary source code repository \url{http://github.com/sunpy/sunpy} can be 
`forked' by anyone with a GitHub account who can then modify their fork as they 
wish. Upon completion of any modifications it is possible for a user to create 
a `Pull Request' (\url{http://github.com/sunpy/sunpy/pulls}) to the main SunPy 
repository in order to incorporate their modification into the SunPy code base. 
Pull requests are reviewed by the SunPy community before being accepted. Community consensus
is required to add new features.
